Addressing climate change and development pressures in an urban estuary through habitat restoration planning

### 2022-12-30 revision/review_response.qmd

* __Response__: Apologies for the omission of that critical detail.  The benchmark period is defined as circa 1950 when the earliest historical aerial photographs of Tampa Bay and its watershed were available.  This time period is defined as pre-development in @Robison10 and the citations that precede the sentence.

    The text was amended as follows: "Priority was given to restoration activities focused on habitat types that were important for a suite of estuarine faunal guilds disproportionately lost or degraded compared to a circa 1950 benchmark period considered as pre-development."

Line 127-128. 5-foot contour is the absolute elevation or elevation relative to sea level?

* __Response__: The explicit definition of the 5-foot contour was previously provided on lines 211-213 as the area of land from the local Mean Lower Low Water elevation extending landward to the elevation 5 feet above Mean Sea Level. The definition was moved to this section for clarity.

@@

Line 140. What are the specific contents of short-term and long-term goals? The reader may want a comparison table of the two goals

* __Response__: The identified short-term targets and long-term goals are provided in Table 5 and they are not presented at this point in the text because the methods that follow describe the process by which they were identified.

     However, we have added some text to provide some clarity on on these terms: "The short-term targets provided an interim set of native habitat coverages to attain within a reasonable planning horizon, after which progress in attaining the long-term goals will be re-assessed."

@@

Line 206-215. The actual available space for restoration efforts depends on the relative relation between wetland elevation and tide height. Whether the slope or elevation of coastal wetlands is used to estimate the potential restoration area under the impact of sea level rise. The author should give specific elevation data and tidal data.

* __Response__: We assume that this comment relates to your general concern that our approach does not provide an explicit quantification of the actual restoration potential of wetlands, either tidal or freshwater.  We certainly agree that identification of wetland restoration areas requires a more detailed assessment of land characteristics, including but not limited to elevation and slope data.  The approach described herein is primarily a screening and prioritization tool that requires additional work for on-site restoration planning. We hope that our response to your comments on lines 304-316 below addresses this concern.

@@

Line 216-220 and Figue2. In addition to the land use/land cover data, I think it is important to consider the digital elevation data as well as the relative sea level rise data.

* __Response__: Please see the response to the previous comment and below regarding comments to lines 304-316.

Line 236. What is the classification standard of xeric, mesic, or hydric soil? Due to sea level rise, wetland soil salinity changes, resulting in the transformation of the spatial pattern of freshwater wetland to saltwater wetland, how to consider this.

* __Response__: As noted in the manuscript, these classifications are defined in @Ries14, but historically are defined in @Baldwin38.  The latter citation was added. Additionally, we certainly agree that soil types may change over time as affected by sea level rise.  However, the coastal stratum included in the 5-foot contour considers the likely rate of sea level rise [sensu @Burke19] within the 2030 and 2050 time period used to establish the targets and goals.  That is, the coastal stratum provides a distinction for the restoration potential of wetlands (mesic/hydric soils) as tidal within the five foot contour (mangroves, salt barrens, salt marshes) and freshwater above the five foot contour (forested and non-forested wetlands) within the expected time period of sea level rise.

     We have added some text at the end of the paragraph to make this clear: "This distinction explicitly accounts for potential salinity changes to soil properties as a function of sea-level rise based on regional projections in the time period for establishing the targets and goals."

@@

Line 248-249. The author needs to specify which data analysis methods were used?

* __Response__: The analysis methods are the set of spatial operations described in the preceding text in the same section and as outlined in Figure 2.  The sentence was revised to make this more clear.  We have also noted two core geospatial functions that were used, as examples, in the text that followed. Readers can view the linked repository for specific details as needed.

@@

Line 304-316. The determination of potential repair area lacks quantitative basis. The priority of potential restoration areas at different locations depends on substrate conditions, ecological importance, engineering implementation, and restoration costs. For the whole region, the restoration potential of various ecosystems should be ranked, and only when they exceed a certain threshold can they be judged as having the operability and real potential of restoration under realistic conditions. Current opportunity maps simply overlap land use type data and soil data. I think such judgment is too subjective and arbitrary, lack of quantitative basis. Since the author mentions that there are data on restoration projects, why not use the projects already carried out to determine whether these potential restoration areas can actually be repaired in the future?

* __Response__: We certainly appreciate the concerns that our approach does not provide a more comprehensive quantitative means of identifying actual restoration potential.  Additional information, as noted by the reviewer, is certainly needed for understanding the true restoration potential of any location within the watershed.  However, we have noted in several locations in the manuscript that the approach described herein is a prioritization tool that can be used as a first step for restoration planning (e.g., lines 83, 90, 113).  Prior to this assessment, comparable tools at the watershed scale were not available and we believe that our approach is valuable in identifying potential restoration areas that regional managers could leverage for planning purposes.  In fact, our use of the term restoration "potential" was meant to describe this tool as a preliminary means to guide follow-up work once sites are identified from the provided opportunity areas.

@@

     That being said, it is worth further clarifying the intent of our approach as a prioritization tool that warrants additional work should restoration projects be pursued at the locations identified by our spatial analyses.  We have added text to two key locations to make this clear.

     First, the following was added to the introduction on line 96: "These opportunity areas provide a first assessment of where restoration could occur, where on the ground assessments could be pursued to further quantify restoration potential."

     Second, the following was added to the methods on line 197: "The identification of these areas on a broad spatial scale serves as a planning tool for restoration practitioners, where follow-up assessments are expected to more fully quantify restoration potential at selected sites."
"
Figure 1. The classification criteria for land use types are confused. I suggest using the Ramsar Convention classification system.

* __Response__: The classification criteria used in the figure was based on the FLUCCS approach described in the text (lines 157), with some categories combined for clarity in the figure. For example, "urban/open lands" includes 13 FLUCCS categories considered relevant for combining and displaying together in the map as a single category to show the extent of development in the Tampa Bay watershed.  The figure caption was revised to describe this approach. An additional citation to @Kawula18 was also added in the text and figure caption. This classification was chosen as this is the standard approach used for the entire state of Florida.

### 2022-12-30 manu-draft.qmd
proof revisions to reviewer 2

In this paper, we describe an approach for habitat restoration and conservation planning that addresses the above challenges by considering the whole watershed, addressing historical changes, focusing on trajectories that have occurred during contemporary time periods, and considering both current and future stressors -- particularly land development and sea-level rise. Current and historical data are available for most Tampa Bay habitats, representing a time period when federal, state and local regulations were in effect and regional impacts from climate change have been documented [@Raabe12; @Cavanaugh14]. The approach establishes a framework that can guide both watershed-level habitat planning and site-level restoration activities and incorporates applicable elements of other habitat restoration paradigms discussed above [@Palmer09]. The general approach includes 1) designation of habitat types by strata relative to the aquatic-terrestrial gradient, 2) quantification of historical trends by habitat types to identify appropriate future targets in coverage, and 3) identification of opportunity areas that could be used by practitioners to achieve restoration goals based on habitat type and past trajectories. These opportunity areas provide a first assessment of where restoration could occur, where on the ground assessments could be pursued to further quantify restoration potential. The outcomes of the approach are spatially specific by providing maps to identity opportunity areas and reproducible using an open science workflow [@Lowndes17] that allows regular updates as new data become available.

@@

Coverage targets for habitat types and opportunity areas for restoration were identified by integrating multiple datasets available for the region. First, habitat status and historical trends were quantified using land use/land cover and subtidal datasets to understand relative changes that have occurred over time. Second, historic habitat restoration efforts conducted in the watershed were synthesized to inform on a practical and feasible level of effort that could be conducted by restoration practitioners in the future. The first two steps were used to identify short-term (2030) targets and long-term (2050) goals for native habitat coverage (hectares). The short-term targets provided an interim set of native habitat coverages to attain within a reasonable planning horizon, at which point progress towards attaining the long-term goals will be re-assessed. Finally, remaining restoration opportunities were spatially identified by combining current coverages with existing or proposed protected areas and areas anticipated to be affected by sea-level rise. As such, the approach identifies reasonable goals and targets based on past trends and provides spatially explicit information that identifies where restoration practitioners could prioritize projects based on opportunities within their respective jurisdictions.

@@

![Land use, cover, and subtidal habitats for the Tampa Bay watershed, Florida, USA. The watershed includes the natural hydrologic boundary with minor modifications to include partners working with the Tampa Bay Estuary Program.  Categories are based on the Florida Land Use Cover and Forms Classification System [@FDOT99; @SWFWMD14; @Kawula18] with relevant codes combined for presentation.](figs/lulcmap.jpg){#fig-lulcmap fig-align="center"}

>>>>

In this paper, we describe an approach for habitat restoration and conservation planning that addresses the above challenges by considering the whole watershed, addressing historical changes, focusing on trajectories that have occurred during contemporary time periods, and considering both current and future stressors -- particularly land development and sea-level rise. Current and historical data are available for most Tampa Bay habitats, representing a time period when federal, state and local regulations were in effect and regional impacts from climate change have been documented [@Raabe12; @Cavanaugh14]. The approach establishes a framework that can guide both watershed-level habitat planning and site-level restoration activities and incorporates applicable elements of other habitat restoration paradigms discussed above [@Palmer09]. The general approach includes 1) designation of habitat types by strata relative to the aquatic-terrestrial gradient, 2) quantification of historical trends by habitat types to identify appropriate future targets in coverage, and 3) identification of opportunity areas that could be used by practitioners to achieve restoration goals based on habitat type and past trajectories. These opportunity areas provide a first assessment of where restoration could occur, where on the ground assessments could be pursued to further quantify restoration potential. The outcomes of the approach are also spatially specific by providing maps to identity opportunity areas and reproducible using an open science workflow [@Lowndes17] that allows regular updates as new data become available.

@@

Coverage targets for habitat types and opportunity areas for restoration were identified by integrating multiple datasets available for the region. First, habitat status and historical trends were quantified using land use/land cover and subtidal datasets to understand relative changes that have occurred over time. Second, historic habitat restoration efforts conducted in the watershed were synthesized to inform on a practical and feasible level of effort that could be conducted by restoration practitioners in the future. The first two steps were used to identify short-term (2030) targets and long-term (2050) goals for native habitat coverage (hectares). The short-term targets provided an interim set of native habitat coverages to attain within a reasonable planning horizon, after which progress in attaining the long-term goals will be re-assessed. Finally, remaining restoration opportunities were spatially identified by combining current coverages with existing or proposed protected areas and areas anticipated to be affected by sea-level rise. As such, the approach identifies reasonable goals and targets based on past trends and provides spatially explicit information that identifies where restoration practitioners could prioritize projects based on opportunities within their respective jurisdictions.

@@

![Land use, cover, and subtidal habitats for the Tampa Bay watershed, Florida, USA. The watershed includes the natural hydrologic boundary with minor modifications to include partners working with the Tampa Bay Estuary Program.  Categories are based on the Florida Land Use Cover and Forms Classification System [@FDOT99; @SWFWMD14; @Kawula18] with relevant codes combined for presentation in the figure.](figs/lulcmap.jpg){#fig-lulcmap fig-align="center"}
